---
title: "Module 4"
author: "Ritika Lama"
date: "2024-10-04"
output: pdf_document
---


**The airquality dataset is a built-in data set in R. Use ?airquality to read the description. We will be interested in predicting the air pollution (Ozone) from the solar radiation (Solar.R), average wind speed (Wind), and maximum daily temperature (Temp). Note that there are several missing values in the airquality data. Create a new data frame to hold airquality[complete.cases(airquality[, c("Ozone", "Solar.R", "Wind", "Temp")]), ] This will remove the rows with missing data from our data frame. Using this cleaned data, answer the following questions.**

```{r data-frame, echo = FALSE}
    data("airquality")
    airquality_data <- airquality[complete.cases(airquality[, c("Ozone", "Solar.R", "Wind", "Temp")]), ]
    head(airquality_data)
```
**1.	Begin by graphically exploring the data. Construct scatterplots of Ozone against Solar.R, Wind, and Temp. You may use either a matrix plot or construct separate plots. Which variables look like they will have linear relationships to the response Ozone? (10 pts)**

```{r scatterplot, echo = FALSE}
library(ggplot2)

# Scatterplot for Ozone vs Solar Radiation
ggplot(airquality_data, aes(x = Solar.R, y = Ozone)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Solar Radiation", y = "Ozone", title = "Ozone vs Solar Radiation")

# Scatterplot for Ozone vs Wind
ggplot(airquality_data, aes(x = Wind, y = Ozone)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Wind", y = "Ozone", title = "Ozone vs Wind")

# Scatterplot for Ozone vs Temperature
ggplot(airquality_data, aes(x = Temp, y = Ozone)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Temperature", y = "Ozone", title = "Ozone vs Temperature")

```

The scatterplots demonstrate that the predictors are each related to the response variable, Ozone, in different ways. Solar Radiation is positively related to Ozone-the higher the Solar Radiation, the higher, typically, the Ozone levels. On the other hand, Wind is negatively related; with increasing winds, Ozone tends to be lower. Lastly, Temperature is also positively linearly related where the higher the temperature level, the higher the Ozone level. Overall, both Solar Radiation and Temperature are positively related with Ozone, while Wind is negatively related; these will indicate that these three variables will be of relevance in modeling the levels of Ozone.

2.	Run a linear regression of Ozone against Solar.R, Wind, and Temp. Produce the coefficients table using the summary() command. Write out the least squares regression equation as a formula. (5 pts)

```{r linear-regression, echo = FALSE}
# Linear regression model
model <- lm(Ozone ~ Solar.R + Wind + Temp, data = airquality_data)

# Summary of the model
summary(model)
```

The least squares regression equation can be written as:
\[
\hat{\text{Ozone}} = -64.34 + 0.05982 \times \hat{\text{Solar.R}} - 3.33359 \times \hat{\text{Wind}} + 1.65209 \times \hat{\text{Temp}}
\]


3.	What does it mean that Wind has a negative coefficient? (5 pts)

The negative coefficient of Wind means that as wind speed increases, Ozone levels are expected to decrease while holding other variables constant.

4.	Referring to your coefficients table from above, which variables are significant at a family-wise error rate of 0.05? Note what values you are comparing. (10 pts)

Looking at the coefficients table, we could determine each variable's significance at the 0.05 family-wise error rate by considering their p-values. The p-value of the intercept is less than 0.05-0.00623, so the intercept is significant. For the Solar Radiation variable, Solar.R, the p-value is below 0.05-0.01124-so it is a significant predictor of Ozone. Wind has a p-value of 1.52e-06, which, being much less than 0.05, is highly statistically significant. Similarly, Temp has a very small p-value of 2.42e-09, hence showing high significance. Thus, all the variables are significant predictors of Ozone at a 0.05 significance level: Solar.R, Wind, and Temp.

5.	Construct a 95% confidence interval for the coefficient of Temp, assuming all other variables are in the model. (5 pts)
```{r ci, echo = FALSE}
# Confidence intervals for all coefficients
confint(model, "Temp", level = 0.95)
```

The 95% confidence interval estimate of Temp's coefficient is approximately from 1.15 to 2.15. Thus, we are 95% confident that the true effect of a one-unit increase in Temp on Ozone levels is actually between 1.15 and 2.15.

6.	Get the ANOVA table from R. (You do not have to add the rows for regression nor total.) What is the value of the MSE and what does it estimate? (10 pts)
```{r anova, echo = FALSE}
# ANOVA table
anova_model <- anova(model)
anova_model
```
From the ANOVA table, the Mean Square Error (MSE) can be calculated as follows:

The degrees of freedom (Df) for the residuals = 107

The sum of squares (Sum Sq) for the residuals = 48003

\[
\text{MSE} = \frac{\text{Sum of Squares of Residuals}}{\text{Degrees of Freedom for Residuals}} = \frac{48003}{107} \approx 448.77
\]

The MSE of about 449 estimates the average squared difference between the observed Ozone values and the values predicted by the regression model. It summarizes how accurate this model is; the smaller the MSE, the closer the model predictions are to the actually observed values, whereas the larger the MSE, the greater the differences between the predicted and the actual values.

7.	Calculate the SSR for this model. (5 pts)
```{r ssr, echo = FALSE}
# Sum of Squares for Regression
SSR <- sum((fitted(model) - mean(airquality_data$Ozone))^2)
SSR
```
The SSR for this model is 73799.12.

8.	At the 0.05 significance level, is there evidence that the model is useful in predicting Ozone? Be sure to mention your F value and p-value explicitly. (5 pts)

There is evidence that the regression model is useful in predicting Ozone at a 0.05 level of significance. This is evident from the overall F-statistic and its associated p-value as shown in the ANOVA table. The model F value amounts to 89.094, while the p-value is 9.509e-16. Since this p-value is well below the 0.05 threshold, we reject the null hypothesis that a model with no predictors fits the data equally well as our model with Solar Radiation (Solar.R), Wind, and Temperature (Temp) as predictors. This is indicative of a large amount of variation in the Ozone being explained by the model.

9.	Determine the R2 and R2ADJ values. Interpret the value of R2 in the context of the prediction of Ozone. (5 pts)
```{r r2-r2adj, echo = FALSE}
# R-squared and Adjusted R-squared
summary(model)$r.squared
summary(model)$adj.r.squared
```

The R2 value of 0.6058946 indicates that around 60.59% of the variation in Ozone levels can be explained by the predictors in the model—Solar Radiation, Wind, and Temperature. This suggests the model fits the data moderately well, capturing a significant portion of the variation in Ozone levels. 

The adjusted R2 value, which is 0.5948449, confirms this interpretation. It accounts for the number of predictors and adjusts the R2 value to prevent overestimating the model's effectiveness. Overall, the model does a reasonable job of predicting Ozone levels.

10.	Use R to get the residuals from this model. You may either use the ordinary residuals or the deleted t residuals. Plot the (ordinary or deleted t) residuals against the fitted values. (5 pts)
```{r residuals, echo = FALSE}
# Residuals
residuals <- resid(model)
fitted_values <- fitted(model)

# Plot residuals against fitted values
plot(fitted_values, residuals, main="Residuals vs Fitted", xlab="Fitted Values", ylab="Residuals")
```


11.	Plot the (ordinary or deleted t) residuals against each of the explanatory variables. Referring to these plots and the plot against fitted values, are there any problems with the assumptions of linearity or constant variation? (10 pts)

```{r residuals-plot, echo = FALSE}
plot(airquality_data$Solar.R, residuals, xlab="Solar.R", ylab="Residuals", main="Residuals vs Solar.R")
plot(airquality_data$Wind, residuals, xlab="Wind", ylab="Residuals", main="Residuals vs Wind")
plot(airquality_data$Temp, residuals, xlab="Temp", ylab="Residuals", main="Residuals vs Temp")
```

Residuals plotted against Solar Radiation and Wind show that most variations are randomly around zero, which would not necessarily suggest a violation of linearity, but there is minor clustering that might indicate the relationship could be nonlinear. In the Wind plot, constant variance maintains, but some high residual outliers may potentially be subject to model specification. Checking the residual plot of temperature is also very important, because apparent patterns in residuals may indicate heteroscedasticity. All in all, the residuals are mainly random.

12.	Check normality of the residuals using a normal probability plot. Do the residuals appear reasonably normal?  (5 pts)

```{r normality, echo = FALSE}
# Normal probability plot
qqnorm(residuals)
qqline(residuals)
```

The Q-Q plot shows that, for the most part, the residuals line up on the 45-degree reference line and thus are near normally distributed. However, there are deviations of lower and upper extremes of points bending downward to the left and upward to the right, respectively, indicating a residual distribution with heavier than normal distribution tails and could suggest outliers or a skewed distribution. Therefore, although the residuals are approximately normal, the deviations in both tails suggest that they are not perfectly normally distributed.

13.	Suppose we are interested in all days with Solar.R=200, Wind=8, and Temp=70. Determine the expected response (Ozone) and the 95% confidence interval for the mean Ozone level at these values. (5 pts)

```{r ci-ozone, echo = FALSE}
# Prediction for Solar.R=200, Wind=8, Temp=70
new_data <- data.frame(Solar.R = 200, Wind = 8, Temp = 70)
predict(model, new_data, interval = "confidence")
```


The expected response (mean Ozone level) is approximately 36.60 at the given values. The 95% confidence interval for the mean Ozone level ranges from 29.59 to 43.61. This interval indicates that, with 95% confidence, the true mean Ozone level falls within this range.

14.	Suppose we have a new day, also with Solar.R=200, Wind=8, and Temp=70. Determine the 95% prediction interval for this new individual at these values. (5 pts)
```{r prediction, echo = FALSE}
# Prediction interval for new individual
predict(model, new_data, interval = "prediction")
```

The 95% prediction interval for the new individual at these values ranges from approximately -5.97 to 79.17. This interval is wider than the confidence interval because it accounts for both the uncertainty in estimating the mean and the variability of individual observations. 

15.	Write 1-3 paragraphs summarizing what you found out about this data. (10 pts)

It can be analyzed from the `airquality` dataset that `Solar.R` and `Temp` are positively related with Ozone, while on the other hand, `Wind` is negatively related. This regression model verifies these results since all three predictors were statistically significant at 0.05 level, as verified from the model, and hence the model explains about 60.59% variability in the level of Ozone, which describes the model fitting to be fairly reasonable. There is still unexplained variability by the predictors included, which suggests that other factors beyond those included as predictors may influence Ozone levels.

Residual analysis shows large residuals follow the normal pattern, but small deviations in the tails argue for outliers or a skewed distribution. There are no serious violations in linearity and constant variance, but some minor patterns combined with outliers may indicate further inspection. Variable transformations or adding interactions will help to better the accuracy of the model. Overall, the model is reasonably fitted but can be further improved.




